---
title: "9: Data Visualization"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## LESSON OBJECTIVES
1. Perform simple data visualizations in the R package `ggplot`
2. Develop skills to adjust aesthetics and layers in graphs
3. Apply a decision tree framework for appropriate graphing methods

## SET UP YOUR DATA ANALYSIS SESSION

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse)

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
PeterPaul.nutrients.gathered <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
PeterPaul.chem.nutrients.summaries <- read.csv("./Data/Processed/NTL-LTER_Lake_Summaries_PeterPaul_Processed.csv")
EPAair <- read.csv("./Data/Processed/EPAair_O3_PM25_NC1718_Processed.csv")

EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")

```

## GGPLOT

ggplot, called from the package `ggplot2`, is a graphing and image generation tool in R. This package is part of tidyverse. While base R has graphing capabilities, ggplot has the capacity for a wider range and more sophisticated options for graphing. 
ggplot has only a few rules: 

* The first line of ggplot code always starts with `ggplot()`
* A data frame must be specified within the `ggplot()` function. Additional datasets can be specified in subsequent layers.
* Aesthetics must be specified, most commonly x and y variables but including others. Aesthetics can be specified in the `ggplot()` function or in subsequent layers.
* Additional layers must be specified to fill the plot. 

### Geoms

Here are some commonly used layers for plotting in ggplot: 

* geom_bar
* geom_histogram
* geom_freqpoly
* geom_boxplot
* geom_violin
* geom_dotplot
* geom_point
* geom_errorbar
* geom_smooth
* geom_line
* geom_area
* geom_abline (plus geom_hline and geom_vline)
* geom_text

### Aesthetics
Here are some commonly used aesthetic types that can be manipulated in ggplot: 

* color
* fill
* shape
* size
* transparency

### Plotting continuous variables over time: Scatterplot
```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
ggplot(EPAair, aes(x = Date, y = Ozone)) + 
  geom_point()

O3plot <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone))
print(O3plot)

# Fix this code
O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone), color = "blue")
print(O3plot2)
#naming what color you want needs to be outside of aesthetics
#color = Ozone, is redundant

# Add additional variables
PMplot <- 
  ggplot(EPAair, aes(x = month, y = PM2.5, shape = as.factor(year), color = Site.Name)) +
  geom_point()
print(PMplot)
#shape of point(circle, triangle, etc) changes based on what year it is. Year should be factor. Color based on Site.Name

# Separate plot with facets
PMplot.faceted <-
  ggplot(EPAair, aes(x = month, y = PM2.5, color = as.factor(year), 
  shape = as.factor(year))) +
  geom_point() +
  facet_wrap(vars(Site.Name), nrow = 3)
print(PMplot.faceted)
#facet_wrap: choose variable Site.Name and organize into 3 rows (rather than columns)

PMplot.faceted2 <-
  ggplot(EPAair, aes(x = month, y = PM2.5)) +
  geom_point() +
  facet_grid(Site.Name ~ year)
print(PMplot.faceted2)
#facet_grid: two facets, grid of graphs separated by site and by year
#tilda represents "by"

# Filter dataset within plot building
O3plot.Blackstone <- 
  ggplot(subset(EPAair, Site.Name == "Blackstone"), aes(x = Date, y = Ozone)) + 
  geom_point() +
  geom_line()
print(O3plot.Blackstone)
#only plot one site by subsetting it.

# Exercise: build your own scatterplots of PeterPaul.chem.nutrients

# 1. 
PPnutrients <- ggplot(PeterPaul.chem.nutrients) +
  geom_point(aes(x=daynum, y=temperature_C, color = year4)) +
  facet_wrap(vars(lakename), nrow = 2)
print(PPnutrients)
# Plot surface temperatures by day of  year. 
# Color your points by year, and facet by lake in two rows.


#2. 
PPnutrients2 <- ggplot(PeterPaul.chem.nutrients, aes(x = sampledate, y=temperature_C, color = depth)) +
  geom_point(size = 0.5)
print(PPnutrients2)
# Plot temperature by date. Color your points by depth.
# Change the size of your point to 0.5


```
### Plotting the relationship between two continuous variables: Scatterplot

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
lightvsDO <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, 
  y = dissolvedOxygen)) +
  geom_point()
print(lightvsDO)

# Adjust axes
lightvsDOfixed <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, 
  y = dissolvedOxygen)) +
  geom_point() +
  xlim(0, 400) +
  ylim(0, 20)
print(lightvsDOfixed)

# Depth in the fields of limnology and oceanography is on a reverse scale
tempvsdepth <- 
  #ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth)) +
  ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse()
print(tempvsdepth)

NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth)) +
  geom_point() +
  geom_smooth(method = lm) +
  geom_abline(aes(slope = 16, intercept = 0), lty=2)
print(NvsP)
#geom_smooth=line of best fit (with CI band). Lm means straight line
#some points don't have measurements for one of the variables,will turn grey
#geom_abline defines line with specifications
#lty=2 is dotted

# Exercise: Plot relationships between air quality measurements

# 1. 
AQIvalues <- 
  ggplot(EPAair, aes(x=Ozone, y = PM2.5, color = Site.Name)) +
  geom_point() +
  geom_smooth(method = lm)
print(AQIvalues)
#if aesthetics are in geom_point, then you need aes again if you want to add geom_smooth.

# Plot AQI values for ozone by PM2.5, colored by site. 
# Add a line of best fit for the linear regression of these variables.

  
```

### Plotting continuous vs. categorical variables

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
 # Barplot + error bars
PeterPaul.nutrient.summaries <- PeterPaul.nutrients.gathered %>%
  group_by(lakename, nutrient) %>%
  summarise(sd = sd(concentration), 
            mean = mean(concentration))
#two groups in group_by, need both in same group_by function

Nutrientplot <- 
  ggplot(PeterPaul.nutrients.gathered) +
  geom_bar(aes(x = lakename, y = concentration, fill = as.factor(nutrient)), # why did we use fill?
           position = "dodge", stat = "summary", fun.y = "mean")             # what's happening here?
print(Nutrientplot)
#dodge separates the bars, fill determines the inside of the bar color

Nutrientplot2 <- 
  ggplot(PeterPaul.nutrient.summaries, aes(x = lakename, y = mean, fill = as.factor(nutrient))) + #
  geom_bar(stat = "identity", 
    position = "dodge") + # what does the stat command do?
  geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd), # how do we specify error bars?
                 position = "dodge")
print(Nutrientplot2)
#bar chart requires a command defining the stat (our summary df is exactly as is, so "identity") and our gathered df says summary, fun.y = mean

# Are there more effective ways to produce summary stats for categories?
#yes, error bars go below 0, doesn't make sense

# Box and whiskers plot
Nutrientplot3 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_boxplot(aes(color = nutrient)) # Why didn't we use "fill"?
print(Nutrientplot3)

# Dot plot
Nutrientplot4 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_dotplot(aes(color = nutrient), binaxis = "y", binwidth = 1, 
               stackdir = "center", position = "dodge") #
print(Nutrientplot4)
#plots every value, if there are two values that are the same, it will plot right next to each other. So the wider it is, the more values are there.

# Violin plot
Nutrientplot5 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_violin(aes(color = nutrient)) #
print(Nutrientplot5)

# Frequency polygons
# Using a tidy dataset
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients) +
  geom_freqpoly(aes(x = tn_ug), color = "black") +
  geom_freqpoly(aes(x = tp_ug), color = "darkblue") +
  geom_freqpoly(aes(x = nh34), color = "darkgray") +
  geom_freqpoly(aes(x = no23), color = "gray") +
  geom_freqpoly(aes(x = po4), color = "blue") 
print(Nutrientplot6)
#wide dataset, have to write each one out

# Using a gathered dataset
Nutrientplot7 <-   
  ggplot(PeterPaul.nutrients.gathered) +
  geom_freqpoly(aes(x = concentration, color = nutrient))
print(Nutrientplot7)
#skinny df, easier to code

# Exercise: Plot distributions of AQI values for EPAair

# 1. 
EPAair_summary <- EPAair %>%
  group_by(Site.Name, year) %>%
  summarise(sd = sd(PM2.5, na.rm = TRUE), mean = mean(PM2.5, na.rm = TRUE))

barAQI <- 
  ggplot(EPAair_summary, aes(x= as.factor(year), y = mean, fill = Site.Name)) +
  geom_bar(position = "dodge", stat = "identity") +
  geom_errorbar(aes(ymin=mean-sd, ymax = mean +sd), position = "dodge")
print(barAQI)
#pay close attention to class of variables
# Create a bar chart plus standard deviation error bars for PM2.5, divided by year. 
# Create separate bars for each site. 

# 2. 
# Create a new plot that better depicts the distribution of PM2.5 concentrations. 
# Divide your graph by year and site.
dotAQI <- 
  ggplot(EPAair_summary, aes(x= as.factor(year), y = mean, fill = Site.Name)) +
  geom_dotplot(aes(color = Site.Name), binaxis = "y", binwidth = 1, 
               stackdir = "center", position = "dodge") 
print(dotAQI)
```
