---
title: "Assignment 2: Coding Basics"
author: "Rachel Bash"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on coding basics in R. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio. 
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
You will need to have the correct software installed to do this (see Software Installation Guide)
Press the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your Assignments folder.
8. After Knitting, please submit the completed exercise (PDF file) to the dropbox in Sakai. Please add your last name into the file name (e.g., "Salk_A02_CodingBasics.pdf") prior to submission.

The completed exercise is due on Thursday, 24 January, 2019 before class begins.

## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
seq1 <- seq(1,100,4) #naming the sequence that goes from 1-100, by 4
mean(seq1) #asking for mean of seq1
median(seq1) #asking for median of seq1
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
namevector <- c("Rachel", "Caroline", "Kat", "Jake")
testvector <- c(80,70,50,40)
passvector <- c(TRUE,TRUE,TRUE,FALSE)

testdataframe <- data.frame(namevector,testvector,passvector)
names(testdataframe) <- c("Name","Score","Passed"); 
testdataframe

```

9. QUESTION: How is this data frame different from a matrix?

> ANSWER: The data frame has multiple types of vectors made up of different elements. There is a character vector, a numerical vector, and logical vector.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
passing.grade <- function(testvector) {
  ifelse(testvector>49,TRUE,FALSE)
  
}

test1 <- passing.grade(60); test1
test2 <- passing.grade(40); test2
test3 <- passing.grade(50); test3


```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> ANSWER: `ifelse` worked for me because I used 49 as the cut-off. If the score was greater than 49 (like in `test1`, then the student passed, prompting the answer TRUE as an output. If the score was 49 or below (`test2`) then FALSE was returned as the output. Honestly, both options would work, it just depends on your preferences on how you like it to look. 
