---
title: "Assignment 8: Time Series Analysis"
author: "Rachel Bash"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on time series analysis.

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio. 
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
You will need to have the correct software installed to do this (see Software Installation Guide)
Press the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your Assignments folder.
8. After Knitting, please submit the completed exercise (PDF file) to the dropbox in Sakai. Please add your last name into the file name (e.g., "Salk_A08_TimeSeries.pdf") prior to submission.

The completed exercise is due on Tuesday, 19 March, 2019 before class begins.

## Brainstorm a project topic
1. Spend 15 minutes brainstorming ideas for a project topic, and look for a dataset if you are choosing your own rather than using a class dataset. Remember your topic choices are due by the end of March, and you should post your choice ASAP to the forum on Sakai.

Question: Did you do this?

> ANSWER: Yes

## Set up your session 
2. Set up your session. Upload the EPA air quality raw dataset for PM2.5 in 2018, and the processed NTL-LTER dataset for nutrients in Peter and Paul lakes. Build a ggplot theme and set it as your default theme. Make sure date variables are set to a date format.

```{r}

getwd()
suppressMessages(library(tidyverse))
#install.packages("lubridate")
suppressMessages(library(lubridate))
#install.packages("nlme")
suppressMessages(library(nlme))
#install.packages("lsmeans")
suppressMessages(library(lsmeans))
#install.packages("multcompView")
suppressMessages(library(multcompView))
suppressMessages(library(trend))

PMair <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv") 
PeterPaul.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv")

#Set Date
PeterPaul.nutrients$sampledate <- as.Date(PeterPaul.nutrients$sampledate, 
                                               format = "%Y-%m-%d")
PMair$Date <- as.Date(PMair$Date, format = "%m/%d/%y")

Rachel_theme <- theme_bw(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(Rachel_theme)

```


## Run a hierarchical (mixed-effects) model

Research question: Do PM2.5 concentrations have a significant trend in 2018?

3. Run a repeated measures ANOVA, with PM2.5 concentrations as the response, Date as a fixed effect, and Site.Name as a random effect. This will allow us to extrapolate PM2.5 concentrations across North Carolina.

3a. Illustrate PM2.5 concentrations by date. Do not split aesthetics by site.

```{r}

#a)
ggplot(PMair, aes(x = Date, y = Daily.Mean.PM2.5.Concentration)) +
  geom_point(color="#7fcdbb") +
  labs(x = "Date", y = "PM2.5 concentration in 2018") +
  scale_x_date(date_breaks = "30 days", date_labels = "%b-%d") +
  theme(axis.text.x = element_text(angle = 45,  hjust = 1))

```

3b. Insert the following line of code into your R chunk. This will eliminate duplicate measurements on single dates for each site.
PM2.5 = PM2.5[order(PM2.5[,'Date'],-PM2.5[,'Site.ID']),]
PM2.5 = PM2.5[!duplicated(PM2.5$Date),]

3c. Determine the temporal autocorrelation in your model. 

3d. Run a mixed effects model. 

```{r}
#b)
PMair = PMair[order(PMair[,'Date'],-PMair[,'Site.ID']),]
PMair = PMair[!duplicated(PMair$Date),]

#c)
PMair.auto <- lme(data= PMair, Daily.Mean.PM2.5.Concentration ~ Date, random = ~1|Site.Name) 
summary(PMair.auto) #intercept = 82.2 and Date= -0.004

PMair.auto

ACF(PMair.auto) #=0.4740630033 is second value which is the degree of autocorrelation at first level

#d)
PMair.mixed <- lme(data= PMair, Daily.Mean.PM2.5.Concentration ~ Date, random = ~1|Site.Name, correlation = corAR1(form = ~ Date|Site.Name, value = 0.474),
                method = "REML")
summary(PMair.mixed)

```

Is there a significant increasing or decreasing trend in PM2.5 concentrations in 2018? 

> ANSWER: No. When you account for the autocorrelation, there is not a significant change in PM2.5 concentrations over the course of 2018 (p-value = 0.36)

3e. Run a fixed effects model with Date as the only explanatory variable. Then test whether the mixed effects model is a better fit than the fixed effect model. 

```{r}
PMair.fixed <- gls(data= PMair, Daily.Mean.PM2.5.Concentration ~ Date, method = "REML")
summary(PMair.fixed)


anova(PMair.mixed, PMair.fixed)


```


Which model is better?

> ANSWER: The mixed linear model (which includes Site Name as a random effects is better) because it has a lower AIC value, and the two models are significantly different from one another as shown by the p-value which is < 0.0001.


## Run a Mann-Kendall test

Research question: Is there a trend in total N surface concentrations in Peter and Paul lakes? 

4. Duplicate the Mann-Kendall test we ran for total P in class, this time with total N for both lakes. Make sure to run a test for changepoints in the datasets (and run a second one if a second change point is likely). 

```{r}
PeterPaul.nutrients.surface <- 
  PeterPaul.nutrients %>%
  select(-lakeid, -depth_id, -comments) %>%
  filter(depth == 0) %>%
  filter(!is.na(tn_ug))
  
# Splitting dataset by lake
Peter.nutrients.surface <- filter(PeterPaul.nutrients.surface, lakename == "Peter Lake")
Paul.nutrients.surface <- filter(PeterPaul.nutrients.surface, lakename == "Paul Lake")

#Mann-Kendall test for Peter
mk.test(Peter.nutrients.surface$tn_ug) #there is definitely a trend over time (p-value < 0.0001)

#Pettitt test for Peter
pettitt.test(Peter.nutrients.surface$tn_ug) #change point at 36

#Re-run separate Mann-Kendall for each change point for Peter
mk.test(Peter.nutrients.surface$tn_ug[1:36]) #no change over time for this section
mk.test(Peter.nutrients.surface$tn_ug[37:98]) #another change point detected within this section because p-value is significant and z-value is positive, run pettitt test again

#Pettitt test for Peter to detect where second change point is
pettitt.test(Peter.nutrients.surface$tn_ug[37:98])

#Another separate Mann-Kendall for each change point section for Peter
mk.test(Peter.nutrients.surface$tn_ug[37:57]) #no change over time for this section
mk.test(Peter.nutrients.surface$tn_ug[58:98]) #no change over time for this section



```


```{r}
#Mann-Kendall test for Paul
mk.test(Paul.nutrients.surface$tn_ug) #no significant trend over time for Paul lake (p-value = 0.73)

#Pettitt test for Paul
pettitt.test(Paul.nutrients.surface$tn_ug) #change point detected at 16 but result is not significant, as expected (p-value = 0.096)


```


What are the results of this test?

> ANSWER: Results for Peter: two change points - one at 36 and one at 57. Beyond line 36, if you split up the data even further, you see no significant positive or negative trend. Results for Paul: no change point or positive/negative trend over time. See annotations in above code for p-values.

5. Generate a graph that illustrates the TN concentrations over time, coloring by lake and adding vertical line(s) representing changepoint(s).

```{r}
ggplot(PeterPaul.nutrients.surface, aes(x = sampledate, y = tn_ug, color = lakename)) +
  geom_point() +
  scale_color_manual(values = c("#7fcdbb", "#253494")) +
  geom_vline(xintercept = as.Date("1993-06-02"), color="#253494", lty = 2) +
  geom_vline(xintercept = as.Date("1994-06-29"), color="#253494", lty = 2) +
  labs(x="Date", y="Total Nitrogen Concentration", color="Lake")
  


```


